---
title: "Influence of Metheorogical Factors on Health"
author: "Alexey Grigorev"
date: "12/03/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
default.par = par()

home.dir = "~/rproj/khb-mortality-analysis/data"
setwd(home.dir)
knitr::opts_knit$set(root.dir=home.dir)
```

```{r, warning=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(tidyr)
library(xts)
```

## Reading the data

```{r}
mort = 
  read.table("SSC_Mo.txt", sep='\t', header=T, dec=',') %>%
  filter(complete.cases(.)) %>%
  select(date=Date.1, mortality=Mo.W.M) %>%
  mutate(date=as.Date(date, format="%d.%m.%Y"))

mort.date = range(mort$date)

temp = 
  read.table("Khb_Temp_Daily.txt", sep=';', header=F, dec=',') %>% 
  filter(complete.cases(.)) %>%
  mutate(date=as.Date(ISOdate(V2, V3, V4))) %>%
  select(date, t.min=V6, t.mean=V8, t.max=V10) %>% 
  filter(date >= mort.date[1] & date <= mort.date[2])

press = read.table('Khb_press_dew_wind.txt', sep='\t', header=T, dec='.')
press = press %>% 
  mutate(date=as.Date(ISOdate(Year, Month, Day)),
         wind.speed=sqrt(uw * uw + vw * vw)) %>%
  select(date, pressure=P0, dew.point=Td, temp=Ta, wind.speed) %>% 
  group_by(date) %>%
  summarise(pressure.ampl=max(pressure, na.rm=T)-min(pressure, na.rm=T),
            pressure=mean(pressure, na.rm=T),
            dew.point.ampl=max(dew.point, na.rm=T)-min(dew.point, na.rm=T),
            dew.point=mean(dew.point, na.rm=T),
            wind.speed.max=max(wind.speed, na.rm=T),
            wind.speed.min=min(wind.speed, na.rm=T),
            wind.speed.ampl=wind.speed.max-wind.speed.min,
            wind.speed=mean(wind.speed, na.rm=T),
            t.min=min(temp, na.rm=T), 
            t.max=max(temp, na.rm=T),
            t.mean=round(mean(temp, na.rm=T), digits=3))

temp2 = press %>%
  select(date, t.min, t.mean, t.max) %>%
  filter(date > max(temp$date) & date <= mort.date[2])

temp = rbind(temp, temp2)

press = press %>%
  select(-c(t.min, t.mean, t.max))

temp = merge(temp, press, by='date')
merged = merge(mort, temp, by='date')
```

Some information about wind vectors: http://wx.gmu.edu/dev/clim301/lectures/wind/wind-uv.html


### Holidays


```{r}
wd = weekdays(merged$date, abbreviate=T)
wd = factor(wd, levels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
levels(wd) = 1:7

holidays = read.table("publicholidays-russia.txt") %>% 
  select(date=V1) %>% 
  mutate(date=as.Date(date, format="%d-%m-%Y"))

merged$is.holiday = merged$date %in% holidays$date
merged$date.dof = wd
merged$is.workday = wd %in% 1:5
```


###Extra Indicators 

Will calculate:

- `RH` - relative humidity, formula taken from http://andrew.rsmas.miami.edu/bmcnoldy/Humidity.html

Indexes for warm periods:

- `ET_M` Effective Temperature by Missenard
  - $\text{ET}_M = T - 0.4 \, (T - 10)\, (1 - \text{RH})$, where $T$ is temperature and $\text{RH}$ is relative humidity
- `ET_S` Effective Temperature by Stedman 
  - $\text{ET}_S = 2.719 + 0.994 \, T + 0.016 \, T_d^2$ where $T_d$ is the dewpoint temperature 

```{r}
merged$RH = exp(17.625 * merged$dew.point / (243.04 + merged$dew.point)) /
            exp(17.625 * merged$t.mean / (243.04 + merged$t.mean))

merged$ET_S = 2.719 + 0.994 * merged$t.mean + 0.016 * merged$dew.point ** 2
merged$ET_M = merged$t.mean - 0.4 * (merged$t.mean - 10) * (1 - merged$RH)
```

Let's compare the effective temperature with the actual temperature

```{r}
temp_comp = merged %>% 
  select(Stedman.ET=ET_S, Missenard.ET=ET_M, Mean.T=t.mean) %>%
  gather(key='key', value='temp')

ggplot(data=temp_comp, aes(x=temp, fill=key)) + 
  geom_density(alpha=0.5) +
  labs(x='Temperature', y='Density') + 
  ggtitle("Comparison of T and ETs")
```

Indexes for cold period

- `WCI_S` Siple:
  - $\text{WCI}_S = (10.45 + 10 * \sqrt{v} - v) \, (33 - T)$
- `WCI_C` Court:
  - $\text{WCI}_C = (9 + 10.9 * \sqrt{v} - v) \, (33 - T)$
- `WCI_H` Hill:
  - $\text{WCI}_H = (0.13 + 0.47 * \sqrt{v} - v) \, (36.6 - T)$


```{r}
merged$WCI_S = (10.45 + 10 * sqrt(merged$wind.speed) - merged$wind.speed) *
               (33 - merged$t.mean)
merged$WCI_C = (9 + 10.9 * sqrt(merged$wind.speed) - merged$wind.speed) *
               (33 - merged$t.mean)
merged$WCI_H = (0.13 + 0.47 * sqrt(merged$wind.speed) - merged$wind.speed) *
               (36.6 - merged$t.mean)

temp_comp = merged %>% 
  select(WCI_S, WCI_C, WCI_H, Mean.T=t.mean) %>%
  gather(key='key', value='temp')

ggplot(data=temp_comp, aes(x=temp, fill=key)) + 
  geom_density(alpha=0.5) +
  labs(x='Temperature', y='Density') + 
  ggtitle("Comparison of T and WCI")

```


## Analysis

```{r}
moving.avg = function(data, window) {
  rollmean(zoo(data), window, fill=list(NA, NULL, NA))
}

ggplot(data=merged, aes(y=t.mean, x=date)) +
  geom_point(shape=20, alpha=0.2) + 
  geom_line(aes(y=moving.avg(t.mean, 15)), color='red')

ggplot(data=merged, aes(y=pressure, x=date)) +
  geom_point(shape=20, alpha=0.2) + 
  geom_line(aes(y=moving.avg(pressure, 15)), color='red')

ggplot(data=merged, aes(y=wind.speed, x=date)) +
  geom_point(shape=20, alpha=0.2)


ggplot(data=merged, aes(y=WCI_S, x=date)) +
  geom_point(shape=20, alpha=0.2)
ggplot(data=merged, aes(y=WCI_C, x=date)) +
  geom_point(shape=20, alpha=0.2)
ggplot(data=merged, aes(y=WCI_H, x=date)) +
  geom_point(shape=20, alpha=0.2)

```

```{r}
ggplot(data=merged, aes(y=mortality, x=date)) +
  geom_point(shape=20, alpha=0.2) + 
  geom_smooth()
```

Seems there's increase in 2002-2006, so probably need to mean-normalize each year

```{r}

year = function(date) {
  as.factor(format(date, "%Y"))
}

m2 = merged %>% 
  group_by(year(date)) %>% 
  mutate(mortality=(mortality - mean(mortality)))

ggplot(data=m2, aes(y=mortality, x=date)) +
  geom_point(shape=20, alpha=0.2) + 
  geom_smooth()

```